---
title: "Applying MultiPLIER"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**C. Savonen 2020, based on J. Taroni's notebook:**

In this notebook we'll prep expression data and apply MultiPLIER/recount2 model.

It has been processed with `SCAN::SCANfast` through 
[refine.bio](https://www.refine.bio/).

## Set up

Set up functions we need. 

```{r}
# magrittr pipe
`%>%` <- dplyr::`%>%`

# need PrepExpressionDF
source(file.path("util", "test_LV_differences.R"))

# need GetNewDataB
source(file.path("util", "plier_util.R"))
```

```{r}
# we need this library to convert from Ensembl gene identifiers to gene symbols
# for use with PLIER
library(org.Hs.eg.db)
```

Function for working with refine.bio-processed data (e.g., uses Ensembl
gene identifiers that must be converted to gene symbols).

```{r}
expression_prep <- function(exprs_df) {
  
  # mapIds default behavior -- select whatever it finds first in the case of
  # 1:many mappings
  gene_symbols_df <- 
    AnnotationDbi::mapIds(org.Hs.eg.db, keys = exprs_df$Gene, 
                          column = "SYMBOL", keytype = "ENSEMBL") %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Ensembl")
  colnames(gene_symbols_df)[2] <- "Symbol"
  
  # tack on the gene symbols
  annot_exprs_df <- gene_symbols_df %>%
    dplyr::inner_join(exprs_df, by = c("Ensembl" = "Gene")) %>%
    dplyr::select(-Ensembl)
  colnames(annot_exprs_df)[1] <- "Gene"
  
  # if there are any duplicate gene symbols, use PrepExpressionDF
  if (any(duplicated(annot_exprs_df$Gene))) {
    agg_exprs_df <- PrepExpressionDF(exprs = annot_exprs_df) %>%
      dplyr::filter(!is.na(Gene))
  } else {
    return(annot_exprs_df %>% dplyr::filter(!is.na(Gene)))
  }
  
}
```

#### Directory setup

```{r}
# directory that holds all the gene expression matrices
exprs_dir <- file.path("data", "expression_data", "by_experiment")
plot_dir <- "plots"
results_dir <- "results"

# Create output folders
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in data

### Expression data

GSE28642

```{r}
exprs_file <- file.path(exprs_dir, "GSE28642", "GSE28642.tsv")
exprs_df <- readr::read_tsv(exprs_file)
```

### PLIER model

```{r}
recount_file <- file.path("data", "recount2_PLIER_data", 
                          "recount_PLIER_model.RDS")
recount_plier <- readr::read_rds(recount_file)
```

## Gene identifier conversion and aggregation

We're going to use the default behavior of `mapIds` in
`expression_prep`, i.e., if there are 1:many mappings, just return the first.

### GSE28642

```{r}
prepped_df <- expression_prep(exprs_df)
```

## MultiPLIER

Once the data have been prepped, apply MulitPLIER model. 
We'll use the short wrapper function below since we have to do this twice.
It is only intended to be used in this environment (e.g., `recount.plier` is
in the global environment) which is why we've placed it here.

```{r}
MultiPLIER <- function(agg_exprs_df, output_file) {
  
  # need a matrix where the gene symbols are row names
  exprs_mat <- agg_exprs_df %>%
    tibble::column_to_rownames("Gene") %>%
    as.matrix()
  
  # apply the MultPLIER model
  b_matrix <- GetNewDataB(exprs.mat = exprs_mat,
                          plier.model = recount_plier)
  
  # save to file!
  saveRDS(b_matrix, output_file)
}
```


```{r}
MultiPLIER(agg_exprs_df = prepped_df,
           output_file = file.path(results_dir, 
                                  "GSE28642_recount2_B.RDS"))
```

### Session Info
```{r}
sessionInfo()
```