---
title: "UMAP Visualization"
author: "ALSF CCDL - Candace Savonen"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

*Purpose*: 

Recommended articles on UMAP:   
- [How UMAP works with visuals](https://umap-learn.readthedocs.io/en/latest/how_umap_works.html)  
- [Original paper on UMAP](https://arxiv.org/abs/1802.03426) by McGinnes, Healy & Melville.  

## 1) Set up
Install [`umap` R package](https://cran.r-project.org/web/packages/umap/vignettes/umap.html) 
if you have not installed it previously. 

```{r}
# Install umap if it isn't in the list of installed packages
if (!("umap" %in% installed.packages())) {
  install.packages("umap")
}
```

```{r}
# Set seed
set.seed(12345)

# Magrittr pipe
`%>%` <- dplyr::`%>%`

# Load ggplot2 for graphing
library(ggplot2)
```

Set up output directories.

```{r}
# Make a results directory if it isn't created yet
if (!dir.exists("results")) {
  dir.create("results")
}
# Make a plots directory if it isn't created yet
if (!dir.exists("plots")) {
  dir.create("plots")
}
```

```{r}
# Metadata file name
metadata_filename <- file.path("data", "HOMO_SAPIENS", "metadata_HOMO_SAPIENS.tsv")

# Data file name
data_filename <- file.path("data", "HOMO_SAPIENS", "HOMO_SAPIENS.tsv")
```

## Import and set up data.
Data downloaded from refine.bio are in tab separated values ("tsv") files and 
include a metadata in a separate tsv file. 
These data set was obtained from [refine.bio](https://www.refine.bio/). 

```{r}
# Read in data tsv file
exprs_df <- readr::read_tsv(data_filename, progress = FALSE) %>% 
  # Make the gene column the rownames so the gene names are out of the way for calculations
  tibble::column_to_rownames('Gene') 

# Print out data
exprs_df
```

Import the metadata for this dataset

```{r}
# Read in metadata tsv file
metadata <- readr::read_tsv(metadata_filename) %>% 
  # Get rid of columns that have only NAs
  dplyr::select(-which(apply(is.na(.), 2, all)))

# Print out metadata so we can get an idea of what we have
metadata
```

## Perform Uniform Manifold Approximation.

```{r}
# Calculate umap
umap_dat <- umap::umap(t(exprs_df))

# Make into dataframe 
umap_df <- data.frame(umap_dat$layout) %>% 
  # Turn samples_ids stored as rownames into column
  tibble::rownames_to_column("refinebio_accession_code") %>%
  # Bring the metadata into this df; match by sample ids
  dplyr::inner_join(metadata, 
                    by = 'refinebio_accession_code')
```

Plot the UMAP coordinates.
Label the data points based on their subgroup.

```{r}
# Make a scatterplot with ggplot2
umap_plot <- ggplot(umap_df, aes(x = X1, y = X2, 
                                 shape = experiment_accession)) +
  geom_point() + 
  theme_classic()

# Print out plot here
umap_plot
```

## Save plot to png.

```{r}
# Save as a png
ggsave(file.path("plots", "umap_scatterplot.png"))
```

## 5) Save UMAP coordinates to tsv file.

```{r}
umap_df %>% as.data.frame() %>% 
  readr::write_tsv(file.path("results", "umap_scores.tsv"))
```

Session info: 

```{r}
# Print session info 
sessionInfo()
```
